Finite-size scaling of free energy in the dimer model on a hexagonal domain

### 2019-02-23 AffineLieAlgebras/MQFT2018/paper.tex
TMPh submitted version

\section*{Introduction}
\label{sec:introduction}
The dimer model appeared as a very simplified model of solutions \cite{Fowler-1937}. The molecules
are represented by the rigid tiles on a lattice. The partition function was computed on various
lattices and with different boundary conditions \cite{doi:10.1080/14786436108243366,
  P.W-1961,kenyon2009lectures}.

Dimer models are the integrable lattice models of statistical physics that are under an active
theoretical~\cite{zj2000,ferrari} and numerical
investigation~\cite{ks2018}. Particular dimer model appears for certain choice
of parameters in the well-known six-vertex model  \cite{pronko2017thesis}. Dimer models are
closely related to the theory of alternating matrices \cite{elkies1992alternating1,elkies1992alternating2}

Specifically, the well-known limit shape phenomenon~\cite{vershik1977kerov} was discovered for dimer
model~\cite{kenyon2006dimers} and a connection with the theory of random matrices was
established~\cite{johansson2002non}.

Study of dimer models on various lattices and domains leads to interesting connections with the
geometry of curved manifolds and with spectra of discrete and continuous Dirac and Laplace operators
\cite{kenyon2002laplacian,kenyon2000asymptotic}. Scaling limit of dimer model is proven to be
described by a Gaussian free-field theory \cite{kenyon2001dominos}, but finite-size corrections were
not considered previously. These corrections are important to close the gap between numerical
simulations and theoretical results.

We consider a particular case of the dimer model on a hexagonal domain of hexagonal lattice, that
can be seen as a ``pile of cubes in the corner''. The energy of configuration is the total number of
cubes. For this particular case we use an exact combinatorial formula for the partition function to
derive the expressions for scaling limit of free energy and first three terms of the finite-size
corrections. We show that the first term is identically zero. The second term has a universal part
that is given by the Euler constant and a geometry-dependent part that is written explicitly in
terms of elementary functions. The third term which encapsulates logarithmic dependence on the mesh
size is connected with the central charge of the effective field theory.

This result is supported by numeric simulations presented in our previous publication
\cite{belov2018finite}.

@@

The weight of a particular configuration is given by the exponent of the volume of all cubes or by a
sum of the height function values:
\begin{equation*}
  \label{eq:10}
  E[conf]=\sum_{i,j} h_{ij}=\mathrm{Volume}
\end{equation*}
We set Boltzman constant equal to 1, then the partition function is 
\begin{equation*}
  \label{eq:14}
  Z=\sum_{conf} e^{-\frac{E[conf]}{T}}=\sum_{conf}q^{\mathrm{Volume}[conf]}, 
\end{equation*}
where $q=\exp\left(-1/T\right)$ .

@@

We are interested in the scaling limit, combined with the thermodynamic limit, when $ T\to \infty,$
and $M,N,K\to \infty$, such that ratios $\frac{M}{T}=a,\quad \frac{N}{T}=b, \quad \frac{K}{T}=c$
remain fixed. In what follows we use $\varepsilon=\frac{1}{T}$, which can be seen as the scale of
the model, e.g. mesh size.

@@

\begin{multline}
  \label{eq:4}
  %\frac{f}{T}
  f=- \sum_{i=1}^{M} \sum_{j=1}^{N} \sum_{k=1}^{K} \frac{1}{V}\left\{
    g(i\varepsilon,j\varepsilon,k\varepsilon)+
    \frac{3}{2}\varepsilon\left[g(i\varepsilon,j\varepsilon,k\varepsilon)+g(i\varepsilon,j\varepsilon,k\varepsilon)^{2}\right]\right.\\
  \left.+\frac{7}{6}\varepsilon^{2}\left[g(i\varepsilon,j\varepsilon,k\varepsilon)+3 g(i\varepsilon,j\varepsilon,k\varepsilon)^{2}+
      2 g(i\varepsilon,j\varepsilon,k\varepsilon)^{3}\right]\right\}  + \mathcal{O}(\varepsilon^{3})
\end{multline}

@@

Taking this into account and using the relation \eqref{eq:21} to express the triple sum
$ \sum_{i=1}^{M} \sum_{j=1}^{N} \sum_{k=1}^{K} \frac{1}{V}    g(i\varepsilon,j\varepsilon,k\varepsilon)$
in \eqref{eq:4} as the integral plus corrections, we obtain
\begin{multline}
\label{eq:23}
 f=-\frac{1}{2(ab+bc+ca)}\left\{\int_{0}^{a} \int_{0}^{b}\int_{0}^{c}\frac{\dx \dy \dz}{e^{x+y+z}-1}+\right.\\
  \left.+\sum_{i=1}^{M}\sum_{j=1}^{N}\sum_{k=1}^{K}\frac{\varepsilon^{5}}{12}\left[g(i\varepsilon,j\varepsilon,k\varepsilon)+3
      g(i\varepsilon,j\varepsilon,k\varepsilon)^{2}+2
      g(i\varepsilon,j\varepsilon,k\varepsilon)^{3}\right]\right\}
\end{multline}
Due to the equations \eqref{eq:6}, the corrections are of the same form as the higher order terms in
$\varepsilon$. The linear in $\varepsilon$ term in \eqref{eq:4} is cancelled by three partial
derivatives in \eqref{eq:21}. The coefficient of the quadratic term is changed by the addition of
second derivatives from \eqref{eq:21} to $\frac{1}{12}$.

@@

The function $\frac{1}{e^{x}-1}$ has a divergence, so we can not apply integral approximation directly. We first need to 
subtract and add $\frac{1}{i\varepsilon}$. The sum over $i$ of this term can be approximated using Euler formula:
\begin{equation}
    \label{eq:24}
    \sum_{i=1}^{M}\frac{1}{i}=\frac{1}{M}\sum_{i=1}^{M}\frac{1}{i/M}=\int_{\frac{1}{M}}^{1}\frac{\dx}{x}+\gamma+O(\varepsilon)=-\ln\varepsilon+\ln a+\gamma+O(\varepsilon)
  \end{equation}
Thus we get the integral
\begin{multline}
  \label{eq:13}
    \frac{\varepsilon^{3}}{12}\left[\sum_{i=1}^{M}\left(     \frac{1}{e^{b+c+i\varepsilon}-1}+
      \frac{1}{1-e^{b+i\varepsilon}}+\frac{1}{1-e^{c+i\varepsilon}}+\frac{1}{e^{i\varepsilon}-1}-\frac{1}{i\varepsilon}\right)-\ln\varepsilon+\ln a+\gamma+\mathcal{O}(\varepsilon^{3})\right]\approx\\
   \frac{\varepsilon^{2}}{12}\left[\int_{0}^{a}\left(     \frac{1}{e^{b+c+x}-1}+
      \frac{1}{1-e^{b+x}}+\frac{1}{1-e^{c+x}}+\frac{1}{e^{x}-1}-\frac{1}{x}\right)\dx-\ln\varepsilon+\ln a+\gamma\right]+\mathcal{O}(\varepsilon^{3})    
\end{multline}
The integral is now taken explicitly, it is equal to
\begin{multline}
  \label{eq:22}
     \frac{\varepsilon^{2}}{12}\left[\int_{0}^{a}\left(     \frac{1}{e^{b+c+x}-1}+
         \frac{1}{1-e^{b+x}}+\frac{1}{1-e^{c+x}}+\frac{1}{e^{x}-1}-\frac{1}{x}\right)\dx-\ln\varepsilon+\ln a+\gamma\right]=\\
     \frac{\varepsilon^{2}}{12}\left[\ln \left(\frac{(e^{a}-1)(e^{b}-1)(e^{c}-1)(e^{a+b+c}-1)}{a (e^{a+b}-1)(e^{b+c}-1)(e^{a+c}-1)}\right)-\ln\varepsilon+\ln a+\gamma\right]
\end{multline}

@@

The expansion  (\ref{eq:3}) can be written as
\begin{equation}
  \label{eq:26}
  f=f_{0}+f_{1}\varepsilon + f_{2}\varepsilon^{2} +f_{3}\varepsilon^{2}\ln\varepsilon,
\end{equation}
where the coefficients are
\begin{equation}
  \label{eq:27}
  f_{0}=-\frac{1}{2(ab+bc+ca)}\int_{0}^{a} \int_{0}^{b}\int_{0}^{c}\frac{\dx \dy \dz}{e^{x+y+z}-1},
\end{equation}
\begin{equation}
  \label{eq:28}
  f_{1}=0,
\end{equation}
\begin{equation}
  \label{eq:29}
  f_{2}=-\frac{1}{2(ab+bc+ca)}\frac{1}{12}\left[\ln\left(\frac{(e^{a}-1)(e^{b}-1)(e^{c}-1)(e^{a+b+c}-1)}{(e^{a+b}-1)(e^{b+c}-1)(e^{a+c}-1)}\right)-
    \gamma\right],
\end{equation}
and
\begin{equation}
  \label{eq:30}
  f_{3}=-\frac{1}{2(ab+bc+ca)}\frac{1}{12}
\end{equation}
In the paper \cite{belov2018finite} we have presented results of Monte Carlo simulations using
Metropolis and Wang-Landau algorithms that support the form of expansion (\ref{eq:26}). In
particular we got $f_{1}=0.04\pm 0.04$ which is consistent with $f_{1}=0$.

In the scaling limit $\varepsilon\to 0$ so called ``limit shape phenomenon''~\cite{kenyon2006dimers}
appears in the dimer model. The areas around the corners of the domain are ``frozen'' with height
function values being fixed. An analytical ``Arctic curve'' delimits frozen regions from the region
where the behavior is described by the effective free field theory
\cite{kenyon2009lectures,kenyon2008height,kenyon2006dimers}.

@@

The third term $f_{2}$ depends only on the shape of the domain through $a,b,c$ with a universal
contribution that is equal to the Euler constant $\gamma$. We conjecture that Euler constant will
appear even for a different choice of Kasteleyn weighting, due to the treatment of logarithmic
divergency. In the future work we will consider coordinate dependent values of $q$ and more complex
geometries, as was done in the paper \cite{okounkov2007random}, to check this suggestion.

@@

Since the non-frozen domain is delimited by a smooth boundary, we have $\chi=1$ and
\begin{equation}
  \label{eq:33}
  f_{3}=\frac{1}{2(ab+bc+ca)} \frac{{\bf c}}{6},
\end{equation}
which suggests the value of central charge ${\bf c}=\frac{1}{2}$. This value is in agreement with the
identification of the dimer model with free fermions in the paper \cite{dijkgraaf2009dimer}.

>>>>

\section*{Introduction}
\label{sec:introduction}
The dimer model appeared in attempt to extend a statistical theory of perfect solutions in chemistry
to the case of liquid mixtures with molecules of two very distinct sizes \cite{Fowler-1937}. The
molecules were represented by the rigid tiles on a lattice and the number of tilings was
approximately estimated. But an exact computation was not accessible at the time.

In 1961, Kasteleyn \cite{P.W-1961}, Temperley and Fisher \cite{doi:10.1080/14786436108243366}
represented the partition function of the dimer model as a Pfaffian of the signed adjacency matrix
(``Kasteleyn matrix'') thus allowing the computation of the number of tilings and of free energy scaling limit.
This result was very elegantly used by Fisher to solve Ising model \cite{fisher1966dimer} and by Fan
and Wu \cite{Fan-1970} to compute free energy for a certain case of eight-vertex model.

Further studies of dimer models revealed the connection to the theory of alternating matrices
\cite{elkies1992alternating1,elkies1992alternating2}. Later, the well-known limit shape
phenomenon~\cite{vershik1977kerov} was discovered for dimer models. First, the ``Arctic circle''
theorem was proven for domino tilings of the domain in the form of ``Aztec diamond''
~\cite{1998math......1068J}. Then similar result was obtained for a hexagonal domain on the
hexagonal lattice~\cite{cohn1998shape}. Soon the connection of these results with the theory of
random matrices was established~\cite{johansson2002non}. The papers
\cite{kenyon2006dimers,kenyon2009lectures} present a detailed exposition of the limit shape
phenomenon in the dimer models.

Dimer models are the integrable lattice models of statistical physics that are now under an active
theoretical~\cite{zj2000,ferrari} and numerical investigation~\cite{ks2018}. Computation of
correlation functions is a common problem for all vertex models \cite{colomo2012approach} as well as
for dimer models. Another problem of great interest is the study of limit shapes in various cases
\cite{borodin2010q,di2018tangent}.

Configurations of dimer model on a hexagonal lattice are in one-to-one correspondence with the
configurations of five-vertex model that appears for certain choice of parameters in the well-known
six-vertex model \cite{kapitonov2012weighted,kapitonov2008five}.

Study of dimer models on various lattices and domains led to interesting connections with the
geometry of curved manifolds and with spectra of discrete and continuous Dirac and Laplace operators
\cite{kenyon2002laplacian,kenyon2000asymptotic}. Scaling limit of dimer model is proven to be
described by a Gaussian free-field theory \cite{kenyon2001dominos}, but finite-size corrections were
not considered previously. These corrections are important to close the gap between numerical
simulations and theoretical results.

We consider a particular case of the dimer model on a hexagonal domain of hexagonal lattice, that
can be seen as a ``pile of cubes in the corner''. The energy of configuration is the total number of
cubes. For this particular case we use an exact combinatorial formula for the partition function to
derive the expressions for scaling limit of free energy and first three terms of the finite-size
corrections. We show that the first term is identically zero. The second term is geometry-dependent
and is written explicitly in terms of elementary functions. The third term which encapsulates
logarithmic dependence on the mesh size is connected with the central charge of the effective field
theory.

This result is supported by numeric simulations presented in our previous publication
\cite{1742-6596-1135-1-012024}.

@@

The weight of a particular configuration is given by the exponent of the volume of all cubes or by a
sum of the height function values:
\begin{equation*}
  \label{eq:10}
  E[conf]=\sum_{i,j} h_{ij}=\mathrm{Volume}
\end{equation*}
We set Boltzman constant equal to 1 and choose system of units in such a way that there is no
coupling constant in the expression for energy. Then the partition function is
\begin{equation*}
  \label{eq:14}
  Z=\sum_{conf} e^{-\frac{E[conf]}{T}}=\sum_{conf}q^{\mathrm{Volume}[conf]}, 
\end{equation*}
where $q=\exp\left(-1/T\right)$ .

@@

We are interested in the scaling limit, combined with the thermodynamic limit, when $ T\to \infty,$
and $M,N,K\to \infty$, such that ratios $\frac{M}{T}=a,\quad \frac{N}{T}=b, \quad \frac{K}{T}=c$
remain fixed. In what follows we use $\varepsilon=\frac{1}{T}$, which can be seen as the scale of
the model, e.g. mesh size due to our choice of the system of units.

@@

\begin{multline}
  \label{eq:4}
  %\frac{f}{T}
  f=- \sum_{i=1}^{M} \sum_{j=1}^{N} \sum_{k=1}^{K} \frac{1}{V}\left\{
    \varepsilon g(i\varepsilon,j\varepsilon,k\varepsilon)+
    \frac{3}{2}\varepsilon^{2}\left[g(i\varepsilon,j\varepsilon,k\varepsilon)+g(i\varepsilon,j\varepsilon,k\varepsilon)^{2}\right]\right.\\
  \left.+\frac{7}{6}\varepsilon^{3}\left[g(i\varepsilon,j\varepsilon,k\varepsilon)+3 g(i\varepsilon,j\varepsilon,k\varepsilon)^{2}+
      2 g(i\varepsilon,j\varepsilon,k\varepsilon)^{3}\right]\right\}  + \mathcal{O}(\varepsilon^{4})
\end{multline}
%% Wrong power of $\varepsilon$ below!!

@@

Taking this into account and using the relation \eqref{eq:21} to express the triple sum
$ \sum_{i=1}^{M} \sum_{j=1}^{N} \sum_{k=1}^{K} \frac{1}{V}    \varepsilon g(i\varepsilon,j\varepsilon,k\varepsilon)$
in \eqref{eq:4} as the integral plus corrections, we obtain
\begin{multline}
\label{eq:23}
 f=-\frac{1}{2(ab+bc+ca)}\left\{\int_{0}^{a} \int_{0}^{b}\int_{0}^{c}\frac{\dx \dy \dz}{e^{x+y+z}-1}+\right.\\
  \left.+\sum_{i=1}^{M}\sum_{j=1}^{N}\sum_{k=1}^{K}\frac{\varepsilon^{5}}{12}\left[g(i\varepsilon,j\varepsilon,k\varepsilon)+3
      g(i\varepsilon,j\varepsilon,k\varepsilon)^{2}+2
      g(i\varepsilon,j\varepsilon,k\varepsilon)^{3}\right]\right\}
\end{multline}
Due to the equations \eqref{eq:6}, the corrections are of the same form as the higher order terms in
$\varepsilon$. The linear in $\varepsilon$ term in \eqref{eq:4} is cancelled by three partial
derivatives in \eqref{eq:21}. The coefficient of the quadratic term is changed by the addition of
second derivatives from \eqref{eq:21} to $\frac{1}{12}$.

@@

The function $\frac{1}{e^{x}-1}$ has a divergence, so we can not apply integral approximation directly. We first need to 
subtract and add $\frac{1}{i\varepsilon}$. The sum over $i$ of this term can be approximated using Euler formula:
\begin{equation}
    \label{eq:24}
    \sum_{i=1}^{M}\frac{1}{i}=\frac{1}{M}\sum_{i=1}^{M}\frac{1}{i/M}=\int_{\frac{1}{M}}^{1}\frac{\dx}{x}+\gamma+O(\varepsilon)=-\ln\varepsilon+\ln a+\gamma+O(\varepsilon)
  \end{equation}
Thus we get the integral %%!!!!!
\begin{multline}
  \label{eq:13}
    \frac{\varepsilon^{3}}{12}\left[\sum_{i=1}^{M}\left(     \frac{1}{e^{b+c+i\varepsilon}-1}+
      \frac{1}{1-e^{b+i\varepsilon}}+\frac{1}{1-e^{c+i\varepsilon}}+\frac{1}{e^{i\varepsilon}-1}-\frac{1}{i\varepsilon}\right)-\ln\varepsilon+\ln a+\gamma+\mathcal{O}(\varepsilon^{3})\right]\approx\\
   \frac{\varepsilon^{2}}{12}\left[\int_{0}^{a}\left(     \frac{1}{e^{b+c+x}-1}+
      \frac{1}{1-e^{b+x}}+\frac{1}{1-e^{c+x}}+\frac{1}{e^{x}-1}-\frac{1}{x}\right)\dx-\ln\varepsilon+\ln a+\gamma\right]+\mathcal{O}(\varepsilon^{3})    
\end{multline}
The integral is now taken explicitly, it is equal to
\begin{multline}
  \label{eq:22}
     \frac{\varepsilon^{2}}{12}\left[\int_{0}^{a}\left(     \frac{1}{e^{b+c+x}-1}+
         \frac{1}{1-e^{b+x}}+\frac{1}{1-e^{c+x}}+\frac{1}{e^{x}-1}-\frac{1}{x}\right)\dx-\ln\varepsilon+\ln a+\gamma\right]=\\
     \frac{\varepsilon^{2}}{12}\left[\ln \left(\frac{(e^{a}-1)(e^{b}-1)(e^{c}-1)(e^{a+b+c}-1)}{a (e^{a+b}-1)(e^{b+c}-1)(e^{a+c}-1)}\right)-\ln\varepsilon+\ln a+\gamma\right]
\end{multline}

@@

The expansion  (\ref{eq:3}) can be written as
\begin{equation}
  \label{eq:26}
  f=f_{0}+f_{1}\varepsilon +f_{2}\varepsilon^{2}\ln\varepsilon+ f_{3}\varepsilon^{2},
\end{equation}
where the coefficients are
\begin{equation}
  \label{eq:27}
  f_{0}=-\frac{1}{2(ab+bc+ca)}\int_{0}^{a} \int_{0}^{b}\int_{0}^{c}\frac{\dx \dy \dz}{e^{x+y+z}-1},
\end{equation}
\begin{equation}
  \label{eq:28}
  f_{1}=0,
\end{equation}
\begin{equation}
  \label{eq:30}
  f_{2}=-\frac{1}{2(ab+bc+ca)}\frac{1}{12},
\end{equation}
and
\begin{equation}
  \label{eq:29}
  f_{3}=-\frac{1}{2(ab+bc+ca)}\frac{1}{12}\left[\ln\left(\frac{(e^{a}-1)(e^{b}-1)(e^{c}-1)(e^{a+b+c}-1)}{(e^{a+b}-1)(e^{b+c}-1)(e^{a+c}-1)}\right)-
    \gamma\right].
\end{equation}

In the paper \cite{1742-6596-1135-1-012024} we have presented results of Monte Carlo simulations using
Metropolis and Wang-Landau algorithms that support the form of expansion (\ref{eq:26}). In
particular we got $f_{1}=0.04\pm 0.04$ which is consistent with $f_{1}=0$.

In the scaling limit $\varepsilon\to 0$ so called ``limit shape phenomenon''~\cite{1998math......1068J,cohn1998shape}
appears in the dimer model. The areas around the corners of the domain are ``frozen'' with height
function values being fixed. An analytical ``Arctic curve'' delimits frozen regions from the region
where the behavior is described by the effective free field theory
\cite{kenyon2009lectures,kenyon2008height,kenyon2006dimers}.

@@

@@

Since the non-frozen domain is delimited by a smooth boundary, we have $\chi=1$ and we interpret
$2(ab+bc+ca)$ as a volume of the domain, thus
\begin{equation}
  \label{eq:33}
  \delta F = \left(2(ab+bc+ca)\right) f_{2}\ln\varepsilon,
\end{equation}
which together with formula \eqref{eq:30} suggests the value of central charge
${\bf c}=\frac{1}{2}$. This value is in agreement with the identification of the dimer model with
free fermions in the paper \cite{dijkgraaf2009dimer}.

The last term $f_{3}$ depends only on the shape of the domain through $a,b,c$ with a universal
contribution that is equal to the Euler constant $\gamma$. We conjecture that Euler constant will
appear even for a different choice of Kasteleyn weighting, due to the treatment of logarithmic
divergency. In the future work we will consider coordinate dependent values of $q$ and more complex
geometries, as was done in the paper \cite{okounkov2007random}, to check this suggestion.
